Transitivity of Subtyping for Intersection Types

### 2020-05-13 papers/intersection-transitivity.tex
starting to revise

\begin{abstract}
  The subtyping relation for intersection type systems traditionally
  employs a transitivity rule (Barendregt et al. 1983), which means
  that the subtyping judgment does not enjoy the subformula property.
  Laurent develops a sequent-style subtyping judgment, without
  transitivity, and proves transitivity via a sequence of six lemmas
  that culminate in cut-elimination (2018). This article presents a
  subtyping judgment, in regular style, that satisfies the subformula
  property, and presents a direct proof of transitivity. Borrowing
  from Laurent's system, the rule for function types is essentially
  the $\beta$-soundness property.  The main lemma required for the
  transitivity proof is one that has been used to prove the inversion
  principle for subtyping of function types. The choice of induction
  principle for the proof of transitivity is subtle: we use
  well-founded induction on the lexicographical ordering of the sum of
  the depths of the first and last type followed by the sum of the
  sizes of the middle and last type. The article concludes with a
  proof that the new subtyping judgment is equivalent to that of
  Barendregt, Coppo, and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle, as a tool for studying normalization in the lambda
calculus~\citep{Coppo:1979aa}. By varying the subtyping rules and atom
types, researchers use intersection type systems to model many
different
calculi~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.
Perhaps the best-known of them is the BCD intersection type system of
\citet{Barendregt:1983aa}. For this article we focus on the BCD
system, following the presentation of \citet{Barendregt:2013aa}.  We
conjecture that our results apply to other intersection type systems
that include the $({\to}{\cap})$ rule for distributing intersection
and function types.

>>>>

\begin{abstract}
  The subtyping rules for intersection type systems traditionally
  employ a transitivity rule (Barendregt et al. 1983), which means
  that subtyping does not enjoy the subformula property.  Laurent
  develops a sequent-style subtyping system, without transitivity, and
  proves transitivity via a sequence of six lemmas that culminate in
  cut-elimination (2018). This article develops a subtyping system in
  regular style that omits transitivity and provides a direct proof of
  transitivity. The new system satisfies something like the subformula
  property: every type occuring in the subtyping derivation of $A <:
  B$ is a subformula of $A$ or $B$, or an intersection of such
  subformulas.  Borrowing from Laurent's system, the rule for function
  types is essentially the $\beta$-soundness property.  The main lemma
  required for the transitivity proof is one that has been used to
  prove the inversion principle for subtyping of function types. The
  choice of induction principle for the proof of transitivity is
  subtle: we use well-founded induction on the lexicographical
  ordering of the sum of the depths of the first and last type
  followed by the sum of the sizes of the middle and last type. The
  article concludes with a proof that the new subtyping judgment is
  equivalent to that of Barendregt, Coppo, and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle, as a tool for studying normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
been used for at least three purposes: to enhance the expressiveness
of type systems
~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
to increase the precision of static
analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
and to express filter models for the denotational semantics of a
variety of lambda calculi
~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.

UNDER CONSTRUCTION

A filter model

\[
   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
\]

Perhaps the best-known  them is the BCD intersection type system of
\citet{Barendregt:1983aa}. For this article we focus on the BCD
system, following the presentation of \citet{Barendregt:2013aa}.  We
conjecture that our results apply to other intersection type systems
that include the $({\to}{\cap})$ rule for distributing intersection
and function types.
